---
title: "LogisticalRegression Homework"
author: "Brian Hallberg"
date: "6/19/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##  Get the data needed

Load the states data.
Grab the labels incase we need them later.

```{r}
library(effects)
library(caTools)

NH11 <- readRDS("dataSets/NatHealth2011.rds")
labs <- attributes(NH11)$labels
```
## Exercise: logistic regression
───────────────────────────────────

### Use the NH11 data set that we loaded earlier.

1. Use glm to conduct a logistic regression to predict ever worked (everwrk) using age (age_p) and marital status (r_maritl).
2. Predict the probability of working for each level of marital status.

Note that the data is not perfectly clean and ready to be modeled. You will need to clean up at least some of the variables before fitting the model.

### Lets see what we have first:

```{r}
str(NH11$everwrk)
levels(NH11$everwrk)
str(NH11$age_p)
str(NH11$r_maritl)
levels(NH11$r_maritl)
```

###  Now clean up 

1. Remove all the NA records for everwrk and put in a new table
2. Reduce everwrk to either yes or no.
3. Code yes/no factor into 1/0 and store it in table

```{r}
# Step 1 - remove NA records
NH11work <- subset(NH11, !is.na(everwrk))
# Step 2 - either Yes or No
NH11work$everwrk <- factor(NH11work$everwrk, levels=c("2 No", "1 Yes"))
# Step 3
NH11work$everwrkb <- ifelse(NH11work$everwrk=="1 Yes", 1, 0)
```
### Now split up the data into train and test sets

```{r}
split <- sample.split(NH11work$everwrkb, SplitRatio = 0.65)
Train = subset(NH11work, split == TRUE)
Test = subset(NH11work, split == FALSE)
```

### Now do the Logistical Regression

```{r}
NH11workLog = glm(everwrkb ~ age_p + r_maritl, data = Train, family = binomial)
summary(NH11workLog)
```
###  Try to make the coefficents make more sense

```{r}
NH11workLog.tab <- coef(summary(NH11workLog))
NH11workLog.tab[,"Estimate"] <- exp(coef(NH11workLog))
NH11workLog.tab
```
### Finally plot all the results

```{r}
plot(allEffects(NH11workLog))
summary(allEffects(NH11workLog))
```

