---
title: "Exploring Data and Statistics"
author: "Brian Hallberg"
date: "6/11/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Open Data Files

We need to load the files we want to work with.

```{r getdata}
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)

eligible = readRDS("eligible.rds")
```
##  A couple of fields will be handy for some plots

Added the following fields

* pas - Number of possible all star years for each player based on start of all star games
* pcas - Percent of possible all star games the player appeared in (note: can be >100% since 4 years had 2 all-star games)

Converted fields to integers

* pas
* timein

```{r addfields}
eligible <- eligible %>%
  mutate(pas = case_when (
    debut > '1933-01-01'       ~ timein,
    final_game <= '1933-01-01' ~ 0, 
    TRUE                       ~ timein - floor(difftime('1933-01-01', debut, units="days")/365)
  )
  ) %>%
  mutate(pas = as.integer(pas)) %>%
  mutate(pcas = allstar/pas)
```

##  Summarize some of what we know now

As we proceed a little bit of summary data is provided. The set of data for players covers the years 1871-2015 and includes **18,846** names.

There are **1741** players that were selected to at least 1 All-star games.

There are **247** players in the Baseball Hall of Fame in Cooperstown New York.  So **14.2%** of the All-stars and just **1.3%** of all players that have played baseball in over 100 years have become Hall of Famers.

Looking at eligible players of the **18,846** there are **2385** eligible for the hall of fame which means that only **12.7%** of players even have a chance to be inducted.

## Including Plots

### Scatter plot of the hall of fame players and the number of all star games they participated in

The first thought was that the number of times a player appears in the all-star game may be a good indicator being elected to the hall of fame.  Make some plots to see if we can find some sort of threshold that is important.  First step is to create a file with just hall of fame players and their appearances in the all-star game.  The first all-star game (at least the data for it) is from 1933 so anyone elected before what will not appear in a game, so we dropped those players.

```{r hofallstar}

hofa <- eligible %>%
  filter(inducted == 'Y') %>%
  filter(final_game > '1933-01-01') %>%
  select(player_id, inducted, allstar, timein, final_game, pcas, pas)

ggplot(hofa, aes(x=allstar, y=player_id)) +
  geom_point()

ggplot(hofa, aes(x=pcas, y=player_id)) +
  geom_point()

ggplot(hofa, aes(x=allstar, fill=factor(timein))) +
  geom_histogram(bins=25)

ggplot(hofa, aes(x=pcas*100, fill=factor(timein))) +
  geom_histogram()

```
##  Now we should look at some data other than all-star

###First thing we do is get summary data for the all stars

* Mean data in all numeric fields
* Min data in all numberic fields

```{r stats}
hof <- eligible %>%
  filter(inducted == 'Y')
meanhof <- eligible %>%
  filter(inducted == 'Y') %>%
  filter(final_game > '1933-01-01') %>%
  summarise_if(is.numeric, mean, na.rm = TRUE) %>%
  select(-starts_with("birth_"), -starts_with("death_"), -weight, -height, -ba)
minhof <- eligible %>%
  filter(inducted == 'Y') %>%
  filter(final_game > '1933-01-01') %>%
  summarise_if(is.numeric, min, na.rm = TRUE) %>%
  select(-starts_with("birth_"), -starts_with("death_"), -weight, -height, -ba)
```


###Next is do hof players have data above the mean on a number of statistics?



