---
title: "Milestone Report"
author: "Brian Hallberg"
date: "June 14, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Introduction
This project will attempt to create a predictive model for identifying players that will become memebrs of the Baseball Hall of Fame using career statistics including All-Star Game appearances as well as possibly hits, home runs, runs, career batting average, runs batted in for non pitchers and wins, strike outs, saves, and games pitched for pitchers.

## Data
The project makes use of a data set called the History of Baseball that contains data from professional baseball covering the period of 1871 to 2015.  The data set includes tables for general player information, All-Star Game appearences, Hall of Fame selections, as well as pitching and hitting yearly statistics.  

There are additional tables that are avaialble as part of the data set, but not used in this project.  The data is available on the Kraggle web side and can be obtained at this here:

* https://www.kaggle.com/seanlahman/the-history-of-baseball

The hall of fame players are identified using a inducted field in the hall_of_fame table.  The project will attempt to use player data and all-start appearances to predict which players were inducted.  Significant tables and fields containing the data this project will use are described below.

### Player Table
The player table contains unique code assigned to each player (player_id), as well as the dates of their debut and final game.  Also used is the first and last name of the player.

### All_star Table
The all_star table contains fields for player_id, year, game for those selected for the game.

### Batting Table
The batting table contains fields for player_id, at bats, runs, hits, home runs, and runs batted in all of which are evaluated.

### Hall_of_fame Table
The hall_of_fame table contains fields for player_id and if they were inducted into the hall that year.  Other fields are not used.

### Pitching Table
The pitching table contains fields for player_id, games pitched, wins, and strikeouts which are evaluated.

## Limitations
Although there are fields in tables for batting average and era, those cannot be averaged over multiple seasons.  There is sufficient data to calculate that for players if needed.  There have been exceptions made to the rules for eligibility for the hall of fame and some of the rules have changed over the years, this process ignores exceptions and uses the current rules.

## Data Cleaning
The raw data was contained in Comma Separated Values (CSV) files, one for each table.  Each file was loaded into a separate data frame using read_csv().  Relevant fields were selected using select(). The example below shows the player and batting table files being read into data frames.

### Read in the data

```{r eval = FALSE}
# read player table into data frame.
player <- read_csv("player.csv")

# read batting table into data frame.  
batting <- read_csv("batting.csv")
```
### Summarize data

THe batting and pitching tables are broken down beyond the player so that a row of the table represents the team and year since a player could have played for more that one team in a year.  I only care about the totals for the players career so summaries of these two tables were created with these statements.

```{r summarize, eval = FALSE}
#  Summarize the batting data first
sumhit <- batting %>%
  group_by(player_id) %>%
  summarise_if(is.numeric, sum, na.rm = TRUE) %>%
  select(-year, -stint)
#  Now get the pitching data
sumpitch <- pitch %>%
  group_by(player_id) %>%
  summarise_if(is.numeric, sum, na.rm = TRUE) %>%
  select(-year, -stint, -baopp, -era)
```

The all_star table has a row that represents a player, year and which all-star game they were voted to play (yes, in 4 years they played 2 all-star games in the year).  I only wanted the number of all-star games each played was voted to play in so another summary was done with this statement.

```{r all_star, eval = FALSE}
sumallstar <- all_star %>%
  select(player_id, league_id, year) %>%
  group_by(player_id) %>%
  count(player_id)
```

Finally the hall_of_fame table shows each year a player was voted on by the baseball writers including the number of votes they received.  I only care if the player was inducted so the following statements were used to reduce all the data to a unique set of players inducted.

```{r hof, eval = FALSE}
sumhof <- hof %>%
  filter(category =="Player") %>%
  select(player_id, inducted) %>%
  filter(inducted == "Y")
```
### Build a final data set of all information by player

Since data was spread across different tables, the corresponding data frames needed to be joined in order to pull the data together into a single data frame.  The statements below shows the joining of the four tables into one data frame used.


```{r final, eval=FALSE}
#  Add player and summarized Hall of Fame data
fullplay <- full_join(player, sumhof, by = "player_id")
#  Add the summarized All-Star data
fullplay <- full_join(fullplay, sumallstar, by = "player_id")
#  Add the summarized batting data
fullplay <- full_join(fullplay, sumhit, by = "player_id")
#  And finally add the summarized pitching data
fullplay <- full_join(fullplay, sumpitch, by = "player_id")
```

### Add fields and update some missing data values

Some fields needed to be derived.  For example, the batting average is calculated.  There are also some missing data that is replaced with values to help with plotting.  Examples are included below.

```{r eval = FALSE}
#  Calculate the batting average
fullplay <- fullplay %>% mutate(ba = h/ab)
#  Replace missing data in specific columns
fullplay <- fullplay %>% mutate_at(vars(inducted), funs(replace(., is.na(.), 'N')))
fullplay <- fullplay %>% mutate_at(vars(allstar), funs(replace(., is.na(.), 0)))
```

### Cleanup some bad column names

The joins of the batting and pitching tables, created some conflicting column names and so they were updated to make some plots more readable and also to help simplify some future code.  These are included below

```{r cleancol, eval = FALSE}
fullplay <- rename(fullplay, allstar = n, rallow = r.y, hrallow = hr.y, bballow = bb.y, hallow = h.y)
fullplay <- rename(fullplay, gpitch = g.y, sopitch = so.y, batterhit = hbp.y)
fullplay <- rename(fullplay, game = g.x, run = r.x, hit = h.x, hr = hr.x, bb = bb.x, so = so.x, hbp = hbp.x)
```
### Reduce to Hall of Fame Eligible Players

In this section, reduce the data set of players to those eligible for induction into the hall of fame. The rules say that the player must have played major league baseball for 10 years and they must be out of the game for 5 years.  So we'll reduce the set of players to those that meet those requirements.

```{r eligiblem, eval = FALSE}
#  Add a column that calulates the years a player was in baseball
eligible <- fullplay %>%
  mutate(timein = floor(difftime(final_game, debut, units="days")/365)) %>% 
  filter(timein > 10) %>%
  filter(final_game <= "2011-01-01")
```

```{r loaddata, echo=FALSE}
library(readr)
library(tidyr)
library(dplyr)

eligible = readRDS("eligible.rds")
```
Some fields were added to determine how many possible all star games a player could have appeared in based on when he played.  Using that and the number of all star games they did play we can get a percent of years in the league that a player was an all-star.  This seemed like it might be a good data value to have.

```{r addfields}
eligible <- eligible %>%
  mutate(pas = case_when (
    debut > '1933-01-01'       ~ timein,
    final_game <= '1933-01-01' ~ 0, 
    TRUE                       ~ timein - floor(difftime('1933-01-01', debut, units="days")/365)
  )
  ) %>%
  mutate(pas = as.integer(pas)) %>%
  mutate(pcas = allstar/pas)
```

\pagebreak

## Initial Findings
In order to identify which data might be useful in predicting hall of fame inductions, some exploratory analysis was performed to compare the data between players that are and are not in the hall of fame.  The results of this analysis which indicate a potentially useful data field are shown below.

### All-Star Game Appearances
So one idea is that the number of game apperances would be a good indicator of a future hall of famer. This scatter plot shows players and whether they were inducted and their all star appearances.  The second one shows the players and the percentage of their career they played it the all-star game.

```{r eligallstar}
library(ggplot2)
ggplot(eligible, aes(x=allstar, y=player_id, color=factor(inducted))) +
  geom_point(size = 3, alpha = 0.5, na.rm=TRUE)

ggplot(eligible, aes(x=pcas*100, y=player_id, color=factor(inducted))) +
  geom_point(size = 3, alpha = 0.5, na.rm=TRUE)
```


\pagebreak

### Data for Batters
Here we do some scatter plots of data for hitters.  We remove the data for pitchers since they were not selected to the hall for their hitting abilities.

```{r echo = FALSE}
p <- ggplot(eligible, aes(y=player_id, color=factor(inducted)), size=1, alpha=0.5)
#p + geom_point(aes(x=hit))
#p + geom_point(aes(x=hr))
#p + geom_point(aes(x=ba))
#p + geom_point(aes(x=rbi))
#p + geom_point(aes(x=run))
#p + geom_point(aes(x=w))
#p + geom_point(aes(x=sopitch))
#p + geom_point(aes(x=gpitch))
```


####  Hits
Scatter plot of Hits versus player color coded by induction status

```{r echo=FALSE}
p + geom_point(aes(x=hit), na.rm=TRUE)
```


####  Home Runs

\pagebreak

### Data for Pitchers

#### Wins


#### Game Appearances



\pagebreak


## Approach
The intent is to use ...  

The predictive quality of the model will be evaluated through cross validation.  This model could be compared to a K-means cluster model using Gower distance for the mix of continuous and categorized data.

The dependent variable will be inducted, and the independent variables will be some combination of ...

The final report will outline the process of how this model is developed, any further data cleaning required, and evaluation.