---
title: "LinearRegression Homework"
author: "Brian Hallberg"
date: "6/19/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##  Get the data needed

Load the states data.
Grab the labels incase we need them later.

```{r}
states.data <- readRDS("dataSets/states.rds") 
states.info <- data.frame(attributes(states.data)[c("names", "var.labels")])
```

## Exercise: least squares regression
────────────────────────────────────────

Use the /states.rds/ data set. Fit a model predicting energy consumed per capita (energy) from the percentage of residents living in metropolitan areas (metro). Be sure to

1. Examine/plot the data before fitting the model
2. Print and interpret the model "summary"
3. "plot" the model to look for deviations from modeling assumptions

Select one or more additional predictors to add to your model and repeat steps 1-3. Is this model significantly better than the model with /metro/ as the only predictor?


```{r}
#  1  - Plot the data
plot(states.data$energy, states.data$metro)

#  2  - Fit the model
EnergyReg <- lm(energy ~ metro, data = states.data)

# Print Summary of model
summary(EnergyReg)

#  3  - look for deviations from model assumptions
par(mar = c(4, 4, 2, 2), mfrow = c(1, 2))
plot(EnergyReg, which = c(1, 2))
```
#  Add an other predictor - College

```{r}
EnergyReg1 <- lm(energy ~ metro + college, data = states.data)
summary(EnergyReg1)
```

Adding college made very little difference, let's try a number and see what we learn

```{r}
EnergyReg2 <- lm(energy ~ metro + density + toxic + green + house + senate, data = states.data)
summary(EnergyReg2)
```

That is better, let's start removing some

```{r}
EnergyReg3 <- lm(energy ~ metro + density + toxic + green + senate, data = states.data)
summary(EnergyReg3)
EnergyReg4 <- lm(energy ~ metro + toxic + green, data = states.data)
summary(EnergyReg4)
```
Looks like toxic and green are the best predictors.


## Exercise: interactions and factors
────────────────────────────────────────

### Use the states data set.

1. Add on to the regression equation that you created in exercise 1 by generating an interaction term and testing the interaction.

2. Try adding region to the model. Are there significant differences across the four regions?

```{r}
EnergyReg5 <- lm(energy ~ metro + toxic + green*income, data = states.data)
summary(EnergyReg5)

#Show the results
coef(summary(EnergyReg5))

# make sure R knows region is categorical
states.data$region <- factor(states.data$region)

EnergyReg6 <- lm(energy ~ metro + toxic + green + region, data = states.data)
summary(EnergyReg6)

coef(summary(EnergyReg6))
```

